Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-01-27 ms.tex
more

A galaxy model is fully specified by four parameters: $\Mbh$,
$\Gamma$, $r_b$, and $\beta$.  We compute models for three different
black hole masses, $\Mbh = 10^6$, $10^7$, $10^8 \Msun$.  The
distribution of $\Gamma$ in the \citet{LauerFaber+:2007a} sample is
bimodal, with a concentration of ``core" galaxies with $\Gamma \lsim
0.3$ and a concentration of ``cusp" galaxies with $\Gamma \gsim 0.7$.
We bracket these possibilities by considering models with $\Gamma=0.8$
and $\Gamma=0.1$.  We fix $\beta = 2$ (a typical value) because our results are most sensitive to the
properties of the gas flow on radial scales $\ll r_b$ and do not depend sensitively on the outer power law slope of the stellar density profile. However, the presence of $\rb$ is necessary for to obtain a converged steady state for some regions of our parameter space.  Depending on the model, we
consider four different values of the $\rb$: 50 pc, 100 pc, 200 pc,
and 400 pc, a range motivated by the observed break radii in
the \citet{LauerFaber+:2007a} sample. For the core galaxies ($\Gamma \le$0.3) in this sample the break radius follows the scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with scatter of
approximately one dex. Most the cusp ($\Gamma \ge0.5$) galaxies have
$\rb$ between 100 and 1000 pc, without a clear trend with $\Mbh$. {\bf AG: In practice when it is clear the rb will not have a
  significant effect on rs it may not be necessary to do runs. For
  instance for Mbh=$10^8$ Msun, gamma=0.8, and vw0=600 km/s the
  stagnation radius changes by only 8\%. So for cusp galaxies with
  vw0$\gsim$600 km/s and Mbh$\lsim 10^8$ Msun it seems like it is a
  safe bet to neglect the effect of the break radius on the
  solution. For low heating rates it my be difficult to converged
  solution for large values of rb=200 and 400 pc.}
% However, the presence of $\rb$ is necessary for to obtain a
% converged steady state for some regions of our parameter space. We
% adopt $\rb=100$ pc for cusp galaxies ($\Gamma=0.8$), while for core
% galaxy models we employ the scaling relation $\rb = $, motivated again
% by the \citet{LauerFaber+:2007a} sample. ({\bf BDM: fill in!}).
% Though simplistic, our results are not sensitive to our assumption
% about the value of $\rb$.

@@

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following a star burst (e.g., \citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the overall mass return rate from evolved stars as a function of B-band luminosity.  Although we give the mass return rate per unit volume, our expressions are equivalent.}, although we will show later that $\eta$ can be higher if star formation is ongoing (Fig.~\ref{fig:eta}).

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy
equation accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random movement
of stars with respect to one another at the characteristic velocity
$\sim \sigma = \sqrt{GM_{\rm enc}/r}$.  This heating is present even for the low
wind velocities $\ll \sigma$ that characterize AGB mass loss in old
stellar populations.  The second term, $v_{w}^{2}$, parameterizes
the contribution of additional sources of heating, such as faster main
sequence or Wolf-Rayet stellar winds, spin-down energy from millisecond pulsars, supernovae, AGN
feedback, etc.  As discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ is uncertain and will in general depend on the SMBH mass and
the age of the stellar population.

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Heat transport by electron conductivity results in an an additional term on the right hand side of the entropy equation of the form  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering becomes comparable to the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field perpendicular to the flow could in principle suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).

@@

Our analytic results above neglect the effects of heat conduction near the stagnation radius, an assumption we now check.  The ratio of the magnitude of conductive heating to the nominal level of ``wind" heating is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}} \tilde{v}_{w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  1.0(\phi/0.1)/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  Because the stagnation radius is usually well inside the break radius, we have approximated the stellar density as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the saturated case, heat conduction is at most an order unity correction for expected values of the saturation parameter $\phi < 0.1$, justifying its neglect in our analysis.

>>>>

A galaxy model is fully specified by four parameters: $\Mbh$,
$\Gamma$, $r_b$, and $\beta$.  We compute models for three different
black hole masses, $\Mbh = 10^6$, $10^7$, $10^8 \Msun$.  The
distribution of $\Gamma$ in the \citet{LauerFaber+:2007a} sample is
bimodal, with a concentration of ``core" galaxies with $\Gamma \lsim
0.3$ and a concentration of ``cusp" galaxies with $\Gamma \gsim 0.7$.
We bracket these possibilities by considering models with $\Gamma=0.8$
and $\Gamma=0.1$.  We fix $\beta = 2$ (a typical value) because our results are most sensitive to the
properties of the gas flow on radial scales $\ll r_b$ and do not depend sensitively on the outer power law slope of the stellar density profile. However, the presence of $\rb$ is necessary for to obtain a converged steady state for some regions of our parameter space.  Depending on the model, we
consider up to four different values of the $\rb$: 50 pc, 100 pc, 200 pc,
and 400 pc, a range motivated by the observed break radii in
the \citet{LauerFaber+:2007a} sample. For the core galaxies ($\Gamma \le$0.3) in this sample the break radius follows the scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with scatter of
approximately one dex. Most the cusp ($\Gamma \ge0.5$) galaxies have
$\rb$ between 100 and 1000 pc, without a clear trend with $\Mbh$. {\bf AG: In practice when it is clear the rb will not have a
  significant effect on rs it may not be necessary to do runs. For
  instance for Mbh=$10^8$ Msun, gamma=0.8, and vw0=600 km/s the
  stagnation radius changes by only 8\%. So for cusp galaxies with
  vw0$\gsim$600 km/s and Mbh$\lsim 10^8$ Msun it seems like it is a
  safe bet to neglect the effect of the break radius on the
  solution. For low heating rates it my be difficult to converged
  solution for large values of rb=200 and 400 pc.}
% However, the presence of $\rb$ is necessary for to obtain a
% converged steady state for some regions of our parameter space. We
% adopt $\rb=100$ pc for cusp galaxies ($\Gamma=0.8$), while for core
% galaxy models we employ the scaling relation $\rb = $, motivated again
% by the \citet{LauerFaber+:2007a} sample. ({\bf BDM: fill in!}).
% Though simplistic, our results are not sensitive to our assumption
% about the value of $\rb$.

@@

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following a star burst (e.g., \citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the overall mass return rate from evolved stars as a function of B-band luminosity.  Although we give the mass return rate per unit volume, our expressions are equivalent.}, although we will show later that $\eta$ can be significantly higher if star formation is ongoing (Fig.~\ref{fig:eta}).

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy
equation accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random movement
of stars with respect to one another at the characteristic velocity
$\sim \sigma \approx \sqrt{GM_{\rm enc}/r}$.  This heating is present even in old stellar populations for which mass and energy injection are dominated by AGB winds with low velocities $\ll \sigma$.  The second term, $v_{w}^{2}$, parameterizes
the contribution of additional sources of heating, such as faster main
sequence or Wolf-Rayet stellar winds, spin-down energy from millisecond pulsars, supernovae, AGN
feedback, etc.  As discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ is uncertain and will in general depend on the SMBH mass and
the age of the stellar population.

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Heat transport by electron conductivity results in an an additional term on the right hand side of the entropy equation of the form  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering becomes comparable to the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field oriented perpendicular to the flow could in principle suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).

@@

Our analytic results neglect the effects of heat conduction, an assumption we now check.  The ratio of the magnitude of the conductive heating rate to the nominal level of wind heating at the stagnation radius is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}} \tilde{v}_{w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  1.0(\phi/0.1)/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  Because the stagnation radius is usually well inside the break radius, we have approximated the stellar density as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the saturated case, heat conduction is at most an order unity correction for expected values of the saturation parameter $\phi < 0.1$, justifying its neglect in our analysis.
